---
title: "MLDS 401: Homework 1"
output: pdf_document
date: "2025-09-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

```{r}
library(readr)
StockBeta <- read_csv("401-data/IBM-Apple-SP500 RR Data.csv", 
                      col_types = cols(`S&P 500` = col_number(), 
                                       IBM = col_number(), 
                                       Apple = col_number()), 
                      skip = 1)
StockBeta <- StockBeta[, -5]
summary(StockBeta)
```

```{r}
plot(StockBeta$IBM ~ StockBeta$Apple, 
     main = "Monthly Return Rate of IBM versus Apple",
     xlab = "Monthly Return Rate (%) of Apple",
     ylab = "Monthly Return Rate (%) of IBM")
```


### (a). Make scatter plots of rates of return of IBM versus S&P 500 and of Apple versus S&P 500 and comment on them.

```{r}
plot(StockBeta$IBM ~ StockBeta$"S&P 500", 
     main = "Monthly Return Rate of IBM versus S&P 500",
     xlab = "Monthly Return Rate (%) of S&P 500",
     ylab = "Monthly Return Rate (%) of IBM")
```

### (b). Calculate the $\beta$'s for IBM and Apple with reference to S&P 500. Comment on the relative magnitudes of the $\beta$'s. Which stock had a higher expected return relative to S&P 500?



### (c). Calculate the sample standard deviations (SDs) of the rates of return for S&P 500, IBM and Apple. Also calculate the correlation matrix. Check that $\beta$ = $rs_y/s_x$ for each stock where $r$ is the correlation coefficient between S&P 500 and the given stock, $s_x$ is the sample SD of S&P 500 and $s_y$ is the sample SD of the given stock.



### (d). Explain based on the statistics calculated how a higher expected return is accompanied by higher volatility of the Apple stock.



## Question 2

```{r}
auto <- read.table("~/Desktop/401-data/auto.txt", head=T)
auto$origin = factor(auto$origin, 1:3, c("US", "Europe", "Japan"))
summary(auto)
```


## Question 3

```{r}
library(psych)
pairs.panels(auto[,c(2:8,1)], stars=T, density=T) # part a
```

```{r}
round(cor(auto[,1:7], use="pair"),4) # part b
```

```{r}
fit = lm(mpg~., auto[,1:8]) # part c
summary(fit)
```

