---
title: "hw10"
output: html_document
date: "2025-12-02"
---

Problem 1
```{r}
library(MASS)
library(dplyr)
library(nnet)
library(e1071)

dat = read.csv("NewsDesert.csv") %>%
  mutate(pub3.2023 = cut(Cpub2023, c(-.5, .5, 1.5, 999), c("0", "1", "2+")))
table(dat$pub3.2023)
set.seed(12345)
train = runif(nrow(dat))<0.8
table(train)

train_data <- dat[train, ]
test_data  <- dat[!train, ]
```

a)
```{r}
fit = multinom(pub3.2023~ age + SES21 + Lpopdens2021 + Lblack2021 + Lhisp2021, data=train_data,
               maxit=1000)
summary(fit)
rbind(
  coef(fit),
  twoOVERone = coef(fit)[2, ] - coef(fit)[1, ]
)

predicted1 <- predict(fit, newdata = test_data)
cm = table(actual = test_data$pub3.2023, predicted = predicted1)
n = sum(cm)
(rowsums = apply(cm, 1, sum))
(colsums = apply(cm, 2, sum))
accuracy = sum(diag(cm)) / n
accuracy
```

b)

```{r}
fit2 = lda(pub3.2023~ age + SES21 + Lpopdens2021 + Lblack2021 + Lhisp2021, data=train_data)
predicted2 <- predict(fit2, newdata = test_data)$class
cm <- table(actual = test_data$pub3.2023, predicted = predicted2)
accuracy <- sum(diag(cm)) / sum(cm)
accuracy
```

c)
```{r}
fit3 = qda(pub3.2023~ age + SES21 + Lpopdens2021 + Lblack2021 + Lhisp2021, data=train_data)
predicted3 <- predict(fit3, newdata = test_data)$class
cm <- table(actual = test_data$pub3.2023, predicted = predicted3)
accuracy <- sum(diag(cm)) / sum(cm)
accuracy
```
d)
```{r}
fitnb = naiveBayes(as.factor(pub3.2023)~ age + SES21 + Lpopdens2021 + Lblack2021 + Lhisp2021, data=train_data)
predictednb <- predict(fitnb, newdata = test_data)$class
cm <- table(actual = test_data$pub3.2023, predicted = predictednb)
accuracy <- sum(diag(cm)) / sum(cm)
accuracy
```
Problem 2
```{r}
dat = expand.grid(factory=c("East", "West"), accident=c("No", "Yes"))
dat$y = c(645,1275, 28,31)
tab = matrix(dat$y, nrow=2,
             dimnames=list(factory=c("East", "West"), accident=c("No", "Yes")))
```
a)
```{r}
p_accident = 59/1979
p_west = 1306/1979
p = p_accident * p_west
p *1979
```
b)
```{r}
chisq.test(tab)$expected
```
c)
d)
e)

Problem 3
a)
```{r}
dat$west <- ifelse(dat$factory == "West", 1, 0)
dat$accident_dummy <- ifelse(dat$accident == "Yes", 1, 0)
glm(y ~ factory + accident, poisson, dat)
```
b)
c)
```{r}
fit2 = glm(y ~ factory*accident, poisson, dat)
summary(fit2)
```
d)
e)
Interaction coefficient is .0288 which is below the 5% significance level 
meaning that the interaction is significant
f)
g)
```{r}
G2 <- deviance(fit) - deviance(fit2)
df <- df.residual(fit) - df.residual(fit2)
pval <- 1 - pchisq(G2, df)

G2 #test statistic
df #degrees of freedom
pval #p value
```
h)
```{r}
#log mi1/mi0 = beta2 + beta3 * w
#east means w = 0
#log odds of accident in the east are .681
exp(.68145)
#west means w = 1
.68145 - 3.13705
exp(-2.4556)
```
i)
j)
```{r}
glm(accident ~ factory, binomial, dat)

logit_df <- data.frame(
  accident_yes = tab[, "Yes"],
  accident_no  = tab[, "No"],
  factory = rownames(tab)
)

# Fit logistic regression using grouped/binomial syntax
fit_logit <- glm(cbind(accident_yes, accident_no) ~ factory,
                 family = binomial, data = logit_df)

summary(fit_logit)
```
Problem 4
```{r}
dat = read.csv("NewsDesert.csv") %>%
  mutate(pub3.2023 = cut(Cpub2023, c(-.5, .5, 1.5, 999), c("0", "1", "2+")))
table(dat$pub3.2023)
```
a)
```{r}
fit4 <- glm(Cpub2023 ~ age+SES21+Lpopdens2021+Lblack2021+Lhisp2021, poisson, data=dat)
summary(fit4)
```
age, socioeconomic status, population density, and percentage hispanic were associated with larger amounts of news organizations
percentage black is associated with a smaller amount of news organizations
no variables are not significant
b)
```{r}
fit5 <- glm(Cpub2023 ~ log(Cpub2018+1)+age+SES21+Lpopdens2021+Lblack2021+Lhisp2021, poisson, data=dat)
summary(fit5)
```

the log number of publications in 2018 was associated with higher number of publications in 2023 as well as socioeconomic status
none of the other variables are significant in this model
c)
```{r}
muhat = fit5$fitted.values # =exp(coef(fit)[1]+coef(fit)[2]*dat$x)
term1 = ifelse(dat$Cpub2023==0, 0, dat$Cpub2023*log(muhat/dat$Cpub2023))
term1

fit6 = multinom(pub3.2023~log(Cpub2018+1)+age+SES21+Lpopdens2021+Lblack2021+Lhisp2021, data=dat,
                maxit=1000)

#Poisson model - estimate P(Y=0)
muhat <- fit5$fitted.values  # fitted lambda for Poisson

#Poisson probability of 0 publications
p0_poisson <- dpois(0, lambda = muhat)

#Multinomial model - estimated P(Y=0)
p0_multinom <- predict(fit6, type = "probs")[, "0"]

library(ggplot2)

#Create a data frame of the probabilities
prob_df <- data.frame(
  P0_Poisson = p0_poisson,
  P0_Multinom = p0_multinom
)

#plot
plot(prob_df$P0_Poisson, prob_df$P0_Multinom,
     xlab = "P(Y=0) Poisson",
     ylab = "P(Y=0) Multinomial",
     main = "Comparison of Poisson vs Multinomial predicted probabilities")
abline(0,1, col="red")  # 45-degree line for reference

#scatterplot matrix
pairs(prob_df)
```