---
title: "Homework 3"
output: pdf_document
date: "2025-10-06"
author: "Arielle Weinstein, Jack Bailey, Junbo (Jacob) Lian, Veronica Lin, Jiashu Huang"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

```{r}
auto <- read.table("~/Desktop/401/hw3/auto.txt", head=T)
```

## (a). The origin variable is categorical, where 1=US, 2=Europe and 3=Japan. Submit a table of the variable (i.e., frequency distribution).

```{r}
auto$origin = factor(auto$origin, 1:3, c("US", "Europe", "Japan"))
table(auto$origin)
```

## (b). Regress mpg on origin, weight and year. Examine the diagnostic plots and comment on which assumptions of the linear model, if any, are violated.

```{r}
fit1 <- lm(mpg ~ origin + weight + year, data=auto)
summary(fit1)
```

```{r}
plot(fit1, which=1)
```

```{r}
plot(fit1, which=2)
```

```{r}
plot(fit1, which=3)
```

```{r}
plot(fit1, which=5)
```

## (c). Regress log(mpg) on origin, log(weight), year and year squared. Examine the diagnostic plots and the summary. Comment on whether the model assumptions are roughly satisfied

```{r}
fit2 <- lm(log(mpg) ~ origin + log(weight) + year + I(year^2), data=auto)
summary(fit2)
```

```{r}
plot(fit2, which=1)
```

```{r}
plot(fit2, which=2)
```

```{r}
plot(fit2, which=3)
```

```{r}
plot(fit2, which=5)
```


## d) Use the results from the previous part to describe the effect of year on log(mpg), i.e., is it U-shaped, inverted-U shaped, or linear? If it is nonlinear, where is the minimum or maximum. Submit a graph showing the effect. For you to think about but not turn in: why would year have this effect?

```{r}
b1 = coef(fit2)["year"]
b2 = coef(fit2)["I(year^2)"]

-b1 / (2 * b2)
```

Since b2>0, it is U-shaped as the parabola opens upward. By taking the derivative and setting it to 0, we find the minimum at around year = 67.

```{r}
library(ggplot2)
ggplot(auto, aes(x = year, y = log(mpg))) + 
  geom_smooth(method = "lm", formula = y ~ x + I(x^2)) + 
  labs(title = "The effect of year on log(mpg)")
```

## (e) What does the coefficient for log(weight) tell you? How is unlogged mpg related to unlogged weight?

```{r}
summary(fit2)
```

```{r}
ggplot(auto, aes(x = weight, y = mpg)) + 
  geom_smooth(method = "lm", formula = y ~ x) + 
  labs(title = "The effect of weight on mpg")
```


# Question 5

## (b). You will now check your work with the auto data set. Regress mpg on year and year squared. Note the coefficients

```{r}
fit3 <- lm(mpg ~ year + I(year^2), data = auto)
summary(fit3)
```

## (c). What is the correlation between year and year squared?

```{r}
cor(auto$year, auto$year^2)
```

## (d). What is the mean of year?

```{r}
mean(auto$year)
```

## (e). What is the correlation between year and year squared after mean centering?

```{r}
year_c  <- auto$year - mean(auto$year)
cor(year_c, I(year_c^2))
```

## (f). To understand why the correlation is reduced, plot year squared against year, and separately centered year squared versus centered year.

```{r}
plot(auto$year, auto$year^2,
     xlab = "year", ylab = "year squared",
     main = paste0("corr = ", round(cor(auto$year, auto$year^2), 3)))
```

```{r}
plot(year_c, year_c^2,
     xlab = "centered year", ylab = "centered year squared",
     main = paste0("corr = ", round(cor(year_c, year_c^2), 3)))
```

## (g). Regress mpg on centered year and centered year squared. Note the coefficients

```{r}
fit4 <- lm(mpg ~ year_c + I(year_c^2), data=auto)
summary(fit4)
```

## (h). Substitute your estimates from the previous part into the expressions you derived in part (a) and show that the equal the estimates from part (b)

```{r}
x_bar <- mean(auto$year)
gamma0 <- coef(fit4)[1]
gamma1 <- coef(fit4)[2]
gamma2 <- coef(fit4)[3]

beta0 <- gamma0 - gamma1 * x_bar + gamma2 * x_bar^2
beta1 <- gamma1 - 2 * gamma2 * x_bar
beta2 <- gamma2

rbind("estimates from part (b)" = coef(fit3),
      "expressions in (a)" = c(beta0, beta1, beta2))
```

# Question 6

## (a). Estimate Model 1 and report the output: \[\log(y + 1) = \beta_0 + \beta_1 \log(x + 1)\]


## (b). Estimate Model 2: \[ \log(y + 1) = \beta_0 + \beta_1 \log(x + 1) + \beta_2 \, tx + e\]


## (c). Based on Model 1, does participation have a significant effect on future spending? Test the appropriate hypothesis at the 5% level and interpret the estimate of \(\beta_2\).


## (d). Using Model 1, post-period spending is how many times greater for those who participate than for those who do not? Note that this question asks about spending and not log spending. Another way to ask this question is: suppose there are two people with identical pre-period spending, but one participates and the other does not. More precisely, if \(y_1\) is the post-contest spending of a participant and \(y_0\) is the post-contest spending of a non-participant, how many times greater is \((y_1 + 1)\) than \((y_0 + 1)\)?


## (e). Why is the magnitude of the tx variable so different in Model 2 (0.050) than in Model 1 (0.244)?


## (f). Now consider Model 2. How do the results from Model 2 change your conclusions about how participation affects future spending. I am looking for you to summarize the key learnings from Model 2 succinctly.


## (g). Why do you think that the word count might affect future spending?


## (h). hat do the results of this analysis suggest the company should do in the future when designing social media contests?

