---
title: "hw9"
output: pdf_document
date: "`r Sys.Date()`"
author: "Arielle Weinstein, Jack Bailey, Junbo (Jacob) Lian, Veronica Lin, Jiashu Huang"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(readr)
library(nnet)
```


# Question 2

```{r}
news <- read.csv("401/hw/hw9/NewsDesert.csv")
```

## (a). Create a variable pub3.2023 that takes three values: 0 newspapers, 1 newspaper, or 2+ newspapers. Submit a frequency distribution (table). 

```{r}
news$pub3.2023 <- cut(
  news$Cpub2023,
  breaks = c(-1, 0, 1, Inf),
  labels = c("0", "1", "2+")
)

table(news$pub3.2023)
```

## (b). Use a multinomial regression to predict pub3.2023 from the demographics. Find the missing logit. Interpret all three logits (0 vs. 1, 1 vs. 2+ and 0 vs. 2+).

```{r}
fit1 <- multinom(factor(pub3.2023, levels = c("2+", "0", "1")) ~ 
                  age + SES21 + Lpopdens2021 + Lblack2021 + Lhisp2021, 
                news)
summary(fit1)
```

**0 vs 1 Logit:** 

```{r}
-apply(coef(fit1), 2, diff)
```

A one unit increase in age will increase the log odds of 1 paper versus 0 by .0183
A one unit increase in socioeconomic status will increase the log odds of 1 paper versus 0 by .0998
A one unit increase in population density will increase the log odds of 1 paper versus 0 by .494
A one unit increase in the percentage of the county that is black will decrease the log odds of 1 paper versus 0 by .0978
A one unit increase in the percentage of the county that is hispanic will decrease the log odds of 1 paper versus 0 by .105
 
  
**0 vs 2+ Logit:**

```{r}
coef(fit1)["0", ]
```

A one unit increase in age will increase the log odds of 2+ papers versus 0 by .00019
A one unit increase in socioeconomic status will increase the log odds of 2+ papers versus 0 by .484
A one unit increase in population density will increase the log odds of 2+ papers versus 0 by .849
A one unit increase in the percentage of the county that is black will decrease the log odds of 2+ papers versus 0 by .349
A one unit increase in the percentage of the county that is hispanic will increase the log odds of 2+ papers versus 0 by .0331
  
  
**1 vs 2+ Logit:**

```{r}
coef(fit1)["1", ]
```

A one unit increase in age will decrease the log odds of 2+ papers versus 1 by .018
A one unit increase in socioeconomic status will increase the log odds of 2+ papers versus 1 by .384
A one unit increase in population density will increase the log odds of 2+ papers versus 1 by .355
A one unit increase in the percentage of the county that is black will decrease the log odds of 2+ papers versus 1 by .251
A one unit increase in the percentage of the county that is hispanic will increase the log odds of 2+ papers versus 1 by .138    

## (c). Use a multinomial regression to predict pub3.2023 from the AR1+ variables. Find the missing logit and interpret the model.

```{r}
fit2 <- multinom(factor(pub3.2023, levels = c("2+", "0", "1")) ~ 
                  log(Cpub2018 + 1) + age + SES21 + Lpopdens2021 + Lblack2021 + 
                  Lhisp2021, news)
summary(fit2)
```
  
**0 vs 1 Logit:** 

```{r}
-apply(coef(fit2), 2, diff)
```

A one unit increase in log number of newspapers published in 2018 will increase the log odds of 1 paper versus 0 by 10.8
A one unit increase in age will increase the log odds of 1 paper versus 0 by .0348
A one unit increase in socioeconomic status will increase the log odds of 1 paper versus 0 by .136
A one unit increase in population density will increase the log odds of 1 paper versus 0 by .268
A one unit increase in the percentage of the county that is black will increase the log odds of 1 paper versus 0 by .0428
A one unit increase in the percentage of the county that is hispanic will decrease the log odds of 1 paper versus 0 by .00951  
  
**0 vs 2+ Logit:**

```{r}
coef(fit2)["0", ]
```

A one unit increase in log number of newspapers published in 2018 will increase the log odds of 2+ papers versus 0 by 21.3
A one unit increase in age will increase the log odds of 2+ papers versus 0 by .0203
A one unit increase in socioeconomic status will increase the log odds of 2+ papers versus 0 by .275
A one unit increase in population density will increase the log odds of 2+ papers versus 0 by .257
A one unit increase in the percentage of the county that is black will increase the log odds of 2+ papers versus 0 by .0217
A one unit increase in the percentage of the county that is hispanic will decrease the log odds of 2+ papers versus 0 by .157
  
  
**1 vs 2+ Logit:**

```{r}
coef(fit2)["1", ]
```

A one unit increase in log number of newspapers published in 2018 will increase the log odds of 2+ papers versus 1 by 10.5
A one unit increase in age will decrease the log odds of 2+ papers versus 1 by .0145
A one unit increase in socioeconomic status will increase the log odds of 2+ papers versus 1 by .138
A one unit increase in population density will decrease the log odds of 2+ papers versus 1 by .0104
A one unit increase in the percentage of the county that is black will decrease the log odds of 2+ papers versus 1 by .0211
A one unit increase in the percentage of the county that is hispanic will decrease the log odds of 2+ papers versus 1 by .148
  

## (d). For your two models, find accuracy; per-class precision, recall and F1; and macro precision, recall and F1. What do you conclude about which classes can be easily distinguished versus those that are more difficult to predict?

```{r}
news$pub3.2023 <- factor(news$pub3.2023, levels = c("2+", "0", "1"))
pred1 <- predict(fit1, type = "class")
pred2 <- predict(fit2, type = "class")
cm1 <- table(pred1, news$pub3.2023)
cm2 <- table(pred2, news$pub3.2023)
```

```{r}
get_metrics <- function(cm, class) {
  
  TP <- cm[class, class]
  FP <- sum(cm[class, ]) - TP
  FN <- sum(cm[, class]) - TP
  TN <- sum(cm) - TP - FP - FN
  
  precision <- TP / (TP + FP)
  recall <- TP / (TP + FN)
  f1 <- 2 * precision * recall / (precision + recall)
  
  return(data.frame(
    Class = class,
    Precision = precision,
    Recall = recall,
    F1 = f1
  ))
}
```

**Demographic Model**

```{r}
get_metrics(cm1, "0")
```

```{r}
get_metrics(cm1, "1")
```

```{r}
get_metrics(cm1, "2+")
```

```{r}
macro_prec1 <- mean(c(
  get_metrics(cm1, "2+")$Precision,
  get_metrics(cm1, "0")$Precision,
  get_metrics(cm1, "1")$Precision
), na.rm = TRUE)

macro_rec1 <- mean(c(
  get_metrics(cm1, "2+")$Recall,
  get_metrics(cm1, "0")$Recall,
  get_metrics(cm1, "1")$Recall
), na.rm = TRUE)

macro_f1_1 <- mean(c(
  get_metrics(cm1, "2+")$F1,
  get_metrics(cm1, "0")$F1,
  get_metrics(cm1, "1")$F1
), na.rm = TRUE)

data.frame(macro_prec1, macro_rec1, macro_f1_1)
```


**AR1+ Model**

```{r}
get_metrics(cm2, "0")
```

```{r}
get_metrics(cm2, "1")
```

```{r}
get_metrics(cm2, "2+")
```

```{r}
macro_prec2 <- mean(c(
  get_metrics(cm2, "2+")$Precision,
  get_metrics(cm2, "0")$Precision,
  get_metrics(cm2, "1")$Precision
), na.rm = TRUE)

macro_rec2 <- mean(c(
  get_metrics(cm2, "2+")$Recall,
  get_metrics(cm2, "0")$Recall,
  get_metrics(cm2, "1")$Recall
), na.rm = TRUE)

macro_f1_2 <- mean(c(
  get_metrics(cm2, "2+")$F1,
  get_metrics(cm2, "0")$F1,
  get_metrics(cm2, "1")$F1
), na.rm = TRUE)

data.frame(macro_prec2, macro_rec2, macro_f1_2)
```

**Interpretation**  

The "0" class is the hardest to predict, as the demographic model never correctly identifies counties with zero newspapers. Even with Model 2, although performance improves substantially, this class remains more difficult to detect than the others.

The other two classes are both moderately well predicted than the "0" class, but neither is fully separable using demographics alone. In Model 1, both classes show only modest precision and recall, and their F1 indicate substantial confusion between them. In Model 2, both classes become much easier to distinguish, with high precision, recall, and F1 for each.
